---
title: "Species demultiplexing experiment3"
author: "Philipp Janßen"
output: html_document
---

## 1) Initial species assignment

### Genotyping & demultiplexing of species (cellSNP/vireo)

```{bash mmul10_demultiplexing}
cd /data/home/EBgrant/scRNA_run3/cellSNP
for SAMPLE in d8_1_1 d8_1_2 d8_2_1 d8_2_2 d16_1 d16_2
do
  sbatch run_cellSNP_vireo_hg38_NoIntron.sh $SAMPLE
done
```

Results of hg38 based demultiplexing: 

```{r hg38_demult}
samples = c("hg38_NoIntron_d8_1_1", "hg38_NoIntron_d8_1_2", "hg38_NoIntron_d8_2_1", "hg38_NoIntron_d8_2_2", "hg38_NoIntron_d16_1", "hg38_NoIntron_d16_2")

donor_list <- list()
for(sample in samples){
  donor_list[[sample]] <- read.table(paste0("/data/home/EBgrant/scRNA_run3/cellSNP/",sample,"/vireo_NoIntron/donor_ids.tsv"), header = T)
}

lapply(donor_list, function(x){table(x$donor_id)})

donor_df <- bind_rows(donor_list, .id = "sample")

# Plotting: 
spec_colors <-        c("#e9ecef","lightgrey", "#B0144F", "#3AA6A6", "#F2A518")
names(spec_colors) <- c("unassigned","doublet","human","cyno/rhesus", "orang")

donor_df %>% 
  mutate(species = ifelse(donor_id == "cynomolgus", "cyno/rhesus",donor_id),
         species = factor(species, levels = c("human", "orang", "cyno/rhesus" ,"doublet", "unassigned"))) %>% 
  
  ggplot(aes(x = sample, fill = species))+
    geom_bar()+
    coord_flip()+
    scale_fill_manual(values = spec_colors)+
    theme_bw()+
    labs(x = "" , y ="number of cells")


# save cell barcodes of cynomolgus assigned cells
for(sample in names(donor_list)){
  cellBC <- donor_list[[sample]] %>% 
    filter(donor_id == "cynomolgus") %>% 
    pull(cell)
  
  write.table(cellBC,paste0("/data/home/EBgrant/scRNA_run3/cellSNP/",sample,"/cyno_cells.txt"), row.names = F, col.names = F, quote = F)
}

```


## 2) Distinguish cyno and rhesus based on mmul10 mapping

Genotype cells from new experiment with variants identified in first experiment:
Input: bam files (cynomolgus assigned cells only) of lanes 1 and 2 
       variants from first experiment (early pool)
Output: genotyping of single cells + unsupervised assingments to donors (vireo)

```{bash test_new}
cd  /opt/anaconda3/bin/
source activate
conda activate vireo

cd /data/home/EBgrant/scRNA_run2/cellSNP/cyno_vs_rhesus/

sbatch --wrap='
#!/bin/bash
#SBATCH --mem=20G
#SBATCH --cpus-per-task=20

workdir=/data/home/EBgrant/scRNA_run2/cellSNP/cyno_vs_rhesus/
name=lane1_cyno_rhesus

# cellSNP input ###
bam=/data/home/EBgrant/scRNA_run2/map/cellranger/lane1_mmul10/outs/possorted_genome_bam.bam
cellBCs=/data/home/EBgrant/scRNA_run2/cellSNP/cyno_vs_rhesus/BC_lane1_cynomolgus.txt
cellSNP_dir=$workdir/$name.cellSNP
REGION_VCF=/data/home/EBgrant/scRNA_run1/cellSNP/create_reference/cyno_rhesus_variants_mmul10_early.cellSNP/cellSNP.base.vcf.gz

# vireo input ###
n_donor=2
OUT_DIR=$workdir/$name.vireo

srun mkdir $cellSNP_dir

srun cellsnp-lite -s $bam -b $cellBCs -O $cellSNP_dir -R $REGION_VCF --minMAF 0 --minCOUNT 5 --cellTAG CB --UMItag UB --gzip

srun vireo -c $cellSNP_dir -N $n_donor -o $OUT_DIR
'

sbatch --wrap='
#!/bin/bash
#SBATCH --mem=20G
#SBATCH --cpus-per-task=20

workdir=/data/home/EBgrant/scRNA_run2/cellSNP/cyno_vs_rhesus/
name=lane2_cyno_rhesus

# cellSNP input ###
bam=/data/home/EBgrant/scRNA_run2/map/cellranger/lane2_mmul10/outs/possorted_genome_bam.bam
cellBCs=/data/home/EBgrant/scRNA_run2/cellSNP/cyno_vs_rhesus/BC_lane2_cynomolgus.txt
cellSNP_dir=$workdir/$name.cellSNP
REGION_VCF=/data/home/EBgrant/scRNA_run1/cellSNP/create_reference/cyno_rhesus_variants_mmul10_early.cellSNP/cellSNP.base.vcf.gz

# vireo input ###
n_donor=2
OUT_DIR=$workdir/$name.vireo

srun mkdir $cellSNP_dir

srun cellsnp-lite -s $bam -b $cellBCs -O $cellSNP_dir -R $REGION_VCF --minMAF 0 --minCOUNT 5 --cellTAG CB --UMItag UB --gzip

srun vireo -c $cellSNP_dir -N $n_donor -o $OUT_DIR
'
```

### Compare donors to reference genome (mmul10)

```{r reference_nucleotide}
library(Rsamtools)
library(BSgenome)

samples = c("mmul10_NoIntron_d8_1_1", "mmul10_NoIntron_d8_1_2", "mmul10_NoIntron_d8_2_1", "mmul10_NoIntron_d8_2_2", "mmul10_NoIntron_d16_1", "mmul10_NoIntron_d16_2")

donor_mmul10_list <- list()
allele_freq_mmul10_list <- list()

for(sample in samples){
  cellSNP_vcf <- vcfR::read.vcfR(paste0("/data/home/EBgrant/scRNA_run3/cellSNP/", sample, "/cellsnp/cellSNP.base.vcf.gz"))
  DP <- Matrix::readMM(paste0("/data/home/EBgrant/scRNA_run3/cellSNP/",sample,"/cellsnp/cellSNP.tag.DP.mtx"))
  AD <- Matrix::readMM(paste0("/data/home/EBgrant/scRNA_run3/cellSNP/",sample,"/cellsnp/cellSNP.tag.AD.mtx"))
  donor_id <- read.table(paste0("/data/home/EBgrant/scRNA_run3/cellSNP/",sample,"/vireo/donor_ids.tsv"), header = T)
  
  colnames(DP) <- donor_id$cell
  colnames(AD) <- donor_id$cell
  
  gt_df <- as.data.frame(cellSNP_vcf@fix) %>% 
    mutate(AD = rowSums(AD), DP = rowSums(DP))
  
  for(donor in grep("donor", unique(donor_id$donor_id), value = T)){
   gt_df[donor] <- rowSums(AD[,donor_id$donor_id == donor]) / rowSums(DP[,donor_id$donor_id == donor])
  }
  
  gt_df_topPos <- gt_df %>% 
    filter(abs(donor1 - donor0) > 0.8) %>% 
    arrange(-DP) %>% 
    dplyr::slice(1:1000)
  
  # Get genome nucleotide
  fasta_file <- FaFile(file="/data/ngs/genomes/refdata-cellranger-6.1.1/mmul10/fasta/genome.fa")
  gr1 <- GRanges(gt_df_topPos$CHROM, IRanges(start=as.numeric(gt_df_topPos$POS), end=as.numeric(gt_df_topPos$POS)))
  refbase <- getSeq(fasta_file, gr1)
  gt_df_topPos$mmul10_allele <- as.data.frame(refbase)$x
  
  gt_df_topPos_long <- gt_df_topPos %>% 
     pivot_longer(cols = grep("donor", colnames(gt_df_topPos), value = T), values_to = "alt_allele_freq", names_to = "donor") %>% 
     mutate(mmul10_allele_freq = case_when(REF == mmul10_allele ~ (1-alt_allele_freq),
                                           ALT == mmul10_allele ~ alt_allele_freq))
  
  
  donor_assignment <- gt_df_topPos_long %>% 
    group_by(donor) %>% 
    dplyr::summarize(median_mmul10_allele_freq = median(mmul10_allele_freq),
              species_mmul10 = ifelse(median_mmul10_allele_freq > 0.7, "rhesus", "cynomolgus"))
  
  donor_id <- left_join(donor_id, donor_assignment, by = c("donor_id" ="donor")) %>% 
    mutate(species_mmul10 = ifelse(!is.na(species_mmul10), species_mmul10, donor_id))
  gt_df_topPos_long <- left_join(gt_df_topPos_long, donor_assignment)
  
  donor_mmul10_list[[sample]] <- donor_id
  allele_freq_mmul10_list[[sample]] <- gt_df_topPos_long
}


lapply(allele_freq_mmul10_list, function(x){x %>% 
  group_by(donor) %>% 
  dplyr::summarize(sum(mmul10_allele_freq > 0.6) / length(mmul10_allele_freq))})
```

```{r plot_allele_freq, fig.width=10, fig.height=6}
library(ggbeeswarm)
species_colors <-        c("#e9ecef","lightgrey", "#B0144F", "#92AEC8", "#3AA6A6", "#F2A518")
names(species_colors) <- c("unassigned","doublet","human","rhesus","cynomolgus", "orang")

plot_mmul10_alleles <- function(x){
  ggplot(x, aes(x = donor, y = mmul10_allele_freq, color = species_mmul10))+
    geom_quasirandom(size = 0.1)+
    scale_color_manual(values = species_colors, limits = force)+
    theme_bw()+
    theme(legend.position = "none")
    #guides(colour = guide_legend(override.aes = list(size=3)))
}

plot_grid(plotlist = lapply(allele_freq_mmul10_list, plot_mmul10_alleles),
          nrow = 2, labels = names(allele_freq_mmul10_list), scale = 0.8)

```

```{r assignment_prop}
donor_df_mmul10 <- bind_rows(donor_mmul10_list, .id = "sample")

donor_df_mmul10 %>% 
  mutate(species_mmul10 = factor(species_mmul10, levels = c("rhesus", "cynomolgus","unassigned", "doublet"))) %>% 
  ggplot(aes(x = sample, fill = species_mmul10))+
      geom_bar()+
      coord_flip()+
      scale_fill_manual(values = species_colors, limits = force)+
      theme_bw()+
      labs(x = "" , y ="number of cells")


donor_colors <-        c("#e9ecef","lightgrey","red", "lightblue")
names(donor_colors) <- c("unassigned","doublet","donor0", "donor1")

donor_df_mmul10 %>% 
  #mutate(species_mmul10 = factor(species_mmul10, levels = c("rhesus", "cynomolgus","unassigned", "doublet"))) %>% 
  ggplot(aes(x = sample, fill = donor_id))+
      geom_bar()+
      coord_flip()+
      scale_fill_manual(values = donor_colors, limits = force)+
      theme_bw()+
      labs(x = "" , y ="number of cells")
```

```{r final_assignments}
species_assignment = list()

for(sample in c("d8_1_1", "d8_1_2", "d8_2_1", "d8_2_2", "d16_1", "d16_2")){
  species_assignment[[sample]] <- donor_list[[paste0("hg38_NoIntron_",sample)]] %>% 
    left_join(select(donor_mmul10_list[[paste0("mmul10_NoIntron_",sample)]], c("cell", "species_mmul10"))) %>% 
    mutate(species = ifelse(donor_id == "cynomolgus", species_mmul10, donor_id)) %>% 
    select(cell, species)
}

species_assignment_df <- bind_rows(species_assignment, .id = "lane") %>% 
  mutate(day = word(lane,1,sep = "_"),
         pool = paste(day,word(lane,2,sep = "_"), sep = "_"))

```

